---
title: "Bad weather damages on health and economy in the USA"
author: "Alessandro Puiatti"
date: "9/22/2017"
output: html_document
---
[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf  "Storm Data Documentation"
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This documet summarizes the data analysis performed over the NOAA Storm Database 
in order to understand which types of events are most harmful with respect to the
population health and whcih have the greatest economic consequences. 
A detailed descritption of the event types and how the harmful and economic
damage are computed can be found at the National Weather Service 
[Storm Data Documentation][1]



## Data Processing
The amount of data reach almost 0.5GB, so it would be better not to load in memory 
the entire table but only the interested ones. For undertsanding which ones they are
we first read 10 lines (the header an some data), just to have an idea of the data
types and to recognize which are the cols that we have to load. In order to
perform this operation we need to load two libraries: *data.table* and *R.utils*.
For further computation other libraries are needed that are: *plyr* and *dplyr*.

```{r}
library(data.table)
library(R.utils,warn.conflicts = FALSE, quietly = TRUE, verbose = FALSE)
library(plyr) # for useful tools
library(dplyr, warn.conflicts = FALSE, quietly = TRUE, verbose = FALSE)

```

Therefore the first opeartions are: unzip the data file (already downloaded), read 
the first 10 lines on a variable (*firstLines*), and then look at them.

```{r}
bunzip2("repdata-data-StormData.csv.bz2", remove = F)
firstLines <- read.csv("repdata-data-StormData.csv", nrows = 10)
firstLines
```

As can be seen from *firstLines* the cols that are of our interest, checking also
their meanings in the [Storm Data Documentation][1], are:

- EVTYPE
- FATALITIES
- INJURIES
- PROPDMG
- PROPDMGEXP
- CROPDMG
- CROPDMGEXP

So the next step si to extract from the database only these 7 cols

```{r}
cols <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", 
          "CROPDMG", "CROPDMGEXP")
data <- fread("repdata-data-StormData.csv", select = cols, showProgress = F)
```

Now, just to have an idea of the memory allocated for *data*

```{r}
sprintf("Size of only seven columns in memory: %s MB", floor(utils::object.size(data)/1000000))
```
that is almost one-tenth of the total memory needed for the entire dataset. 
Now, we check for NAs values
```{r}
mean(is.na(data))
```
Given that there are no NAs we don't need to bother anymore about it in the next steps

The last operations needed for computing the summaries required are then:

1. reorganize the data grouped by *EVTYPE* for answering the question related to the population 
health (*dataHealth*), after havin selcted only the related cols from *data*

2. reorganize the data grouped by *EVTYPE*, for answering the question related to the most 
damageful event (*dataDmg*), after havin selcted only the related cols from *data*


#### Population health

Select the related cols and remove all rows with null values

```{r}
cols <- c("EVTYPE", "FATALITIES", "INJURIES")
dataHealth <- select(data, cols) %>%
        filter(FATALITIES != 0 & INJURIES != 0)
```

Group the data by *EVTYPE* and summarize

```{r}
sumHealth <- group_by(dataHealth, EVTYPE) %>%
        summarize(sumFatal = sum(FATALITIES), sumInjury = sum(INJURIES ))
```  


#### Damges

Select the related cols and remove all rows with null values

```{r}
cols <- c("EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
dataDmg <- select(data, cols) %>%
        filter(PROPDMG != 0 & CROPDMG != 0)
```

The two cols *PROPDMGEXP* and *CROPDMGEXP* have to be converted in integers with the values
equivalent to the amount of expressed unit (look at [Storm Data Documentation][1]):

- 10^3 for *K* 
- 10^6 for *M*
- 10^9 for *B* 

However, looking at the chars presented in the two mention cols, we have other values than
*K*, *M*, and *B*:
```{r}
unique(dataDmg$PROPDMGEXP)
unique(dataDmg$CROPDMGEXP)
```
We can assume that the number "5", "3" and "0" are the power of 10, "m" is equal to "M"
and "k" is equal to "K". The "" element would be considered as NA

```{r}
dataDmg$PROPDMGEXP <- as.numeric(revalue(dataDmg$PROPDMGEXP, c("K"=10^3, "M"=10^6, "m"=10^6,
                                                    "B"=10^9, "5"=10^5, "0"=1, "3"=10^3)))
dataDmg$CROPDMGEXP <- as.numeric(revalue(dataDmg$CROPDMGEXP, c("K"=10^3, "k"=10^3, "M"=10^6, "m"=10^6,
                                                    "B"=10^9, "0"=1)))
```

Then we can create a new table with the *PROPDMG* and *CROPDMG* computed in dollars

```{r}
dataDmg <- transmute(dataDmg, EVTYPE, PROPDMG = PROPDMG*PROPDMGEXP, CROPDMG = CROPDMG*CROPDMGEXP)
```

Group the data by *EVTYPE* and summarize

```{r}
sumDmg <- group_by(dataDmg, EVTYPE) %>%
       summarize( sumCrop = sum(CROPDMG, na.rm = T), sumProp = sum(PROPDMG,na.rm = T))
```  


## Results





